Exploiting First-Class Arrays in Fortran for Accelerator Programming

### 2010-12-14 opencl/papers/intro.tex
text massaging

1. Goal is to write once, transform many. Otherwise potentially reprogram for every architecture.
2. Goal is to code for readability and maintainability, not performance
3. This goal requires expression in a high-level language.
4. But language must be simple enough for compiler to analyze.
5. Thus ideal if language maps well to accelerator architectures.
6. Data parallel constructs in Fortran 90 are chosen.
7. Up to 65 times speed up measured on automatically transformed code.

At Los Alamos National Laboratory (LANL) we have a wide variety of computer
platforms from which to choose.  The common platform is made up of clusters of
compute nodes with standard multi-core processors.  Some nodes also have
accelerators that range from the IBM Cell processor (Roadrunner) to a variety
of GPUs from NVIDIA and AMD.  Some nodes have hardware with vector
instructions and others do not. The peak
performance of the accelerated nodes range from 200+ GFlops for the IBM Cell/B.E.
processor to XXX for NVIDIA Fermi. %% Fermi has another name

%% NEW - CER
However, the performance that these accelerator architectures offer comes at a
cost as processor architectures are trending toward multiple cores with
instances of integrated accelerator units and less of a reliance on memory
caches.  These changes place a heavy burden on application programmers as they
are forced to adapt to the new systems.  An especially challenging problem
faced by application programmers is not only how to program to these new
architectures (considering the massive scale of concurrency available), but
also how to design programs that are portable across the changing landscape of
computer architectures.  The fundamental question we address is what
programming model and language constructs are best suited for these platforms.

@@

A data parallel programming model is well suited to emerging
architectures that are based on either vector processing or massively
parallel collections of simple cores.  A popular trend for programming
systems such as GPU accelerators is the CUDA language from NVidia, or
new languages like Chapel from Cray.  A less intrusive approach that
is well suited to legacy applications and languages are language
extensions or libraries that allow programmers to avoid adopting
entirely new languages.  The recent OpenCL specification is intended
to support this programming model, as are directive-based methods such
as OpenMP or the Accelerator programming model from the Portland
Group~\cite{pgi10accelerator}.

>>>>

In the ideal situation, application programmers would like to adopt
a programming model in which they write their application once and
use automated tools to retarget it to many architectures.  This has
proven to be very challenging historically due to the subtle balance
between high-level expressiveness of code and the performance of the
lower-level code that is emitted by a compiler.  This ideal high-level
model that programmers work with should emphasize readability,
maintainability, and close proximity in abstraction to the problem being
solved.  It should not be corrupted with details of specific target
architectures solely for the purpose of single-system performance.
For certain classes of applications, specifically those that map well
onto a data-parallel programming model, we show that Fortran 90 contains
language features that encourage high-level programming abstractions
without sacrificing performance during low-level code generation.

%1. Goal is to write once, transform many. Otherwise potentially reprogram for every architecture.
%2. Goal is to code for readability and maintainability, not performance
%3. This goal requires expression in a high-level language.
%4. But language must be simple enough for compiler to analyze.
%5. Thus ideal if language maps well to accelerator architectures.
%6. Data parallel constructs in Fortran 90 are chosen.
%7. Up to 65 times speed up measured on automatically transformed code.

At Los Alamos National Laboratory (LANL), as with many supercomputing
facilities today, users have a wide variety of computer platforms from
which to choose.  The most common platform is made up of clusters of
compute nodes with standard multi-core processors.  An increasingly
common feature is that some nodes also have accelerators that range
from the IBM Cell processor (such as the LANL Roadrunner system) to a
variety of GPUs from NVIDIA and AMD.  Some nodes have hardware with
vector instructions and others do not.  The peak performance of these
accelerated nodes often resides in the hundreds of gigaflops.

%The peak performance of the
%accelerated nodes range from 200+ GFlops for the IBM Cell/B.E.
%processor to XXX for NVIDIA Fermi. %% Fermi has another name

However, the performance that these accelerator architectures offer
comes at a cost as processor architectures are trending toward
multiple cores with instances of integrated accelerator units with
user managed memory and less of a reliance on superscalar instruction
level parallelism and hardware managed memory hierarchies (such as
traditional caches).  These changes place a heavy burden on
application programmers as they are forced to adapt to the new
systems.  An especially challenging problem faced by application
programmers is not only how to program to these new architectures
(considering the massive scale of concurrency available), but also how
to design programs that are portable across the changing landscape of
computer architectures with unique memory systems and programming
models.  The fundamental question we address is what programming model
and language constructs are best suited to span this set of new
hardware designs.

@@

A common theme amongst new processors is the emphasis on data parallel
programming.  This model is well suited to emerging architectures that
are based on either vector processing or massively parallel
collections of simple cores.  A popular trend for programming systems
such as GPU accelerators is the CUDA language from NVidia, or new
languages like Chapel from Cray.  A less intrusive approach that is
well suited to legacy applications and languages are language
extensions or libraries that allow programmers to avoid adopting
entirely new languages.  The recent OpenCL specification is intended
to support this programming model, as are directive-based methods such
as OpenMP or the Accelerator programming model from the Portland
Group~\cite{pgi10accelerator}.
